Adaptive broadcast tree for multicores

### 2013-04-29 paper.tex
Ideas on Octopus

\begin{enumerate}
\item Performance of distributed algorithms on multicores. Programmers
  don't get it right due to different complexity measures. We deal
  with this by improving these algorithms. This will lead to higher
  performance of existing algorithms.
\item Diversity of hardware. Hand-tuning algorithms is tedious. Need
  to do this automatically. To evaluate this, we compare the average
  performance of our automatic configuration on a wide range of
  multicores with performance of a single hand-tuned implementation.
\end{enumerate}
We deal with the latter one for now.

% 
Atomic broadcast is a building-block for many higher-level distributed
algorithm such as state-machine replication. Such distributed
algorithms are increasingly used to overcome scalability challenges on
multicore machines. Examples are databases~\cite{Salomie2011,
  Wiesmann2000} and operating systems~\cite{fos:osr09, tornado:osdi99,
  barrelfish:sosp09}.

@@

% Related work
\stefan{Fill in related work: in a nutshell, they do tuning, but only
  for specific machines}

@@

A multicore aware broadcast has been implemented in~\cite{Tu2008}. It
takes the cache architecture of the machine into consideration for
building the broadcast. It is however for larger messages (starting at
4K) and it does not seem to be performing very well (numbers are given
in microseconds -6). 4K messages are 52 microseconds, which is 150000
cycles. For synchronization primitives, a cache line is sufficient.

@@

\subsection{Aggregation}

Aggregation is the act of collecting information from nodes (i.e.\ it
is kind of the reverse of a broadcast). In node on the return path,
information can be aggregated. One example is batching, where several
pieces of information are grouped in the same message to reduce packet
processing overhead. Another example is pre-processing of data. One
example is pre-calculating the average value of child messages if the
sink node is not interested in individual values.

Similarly to what has been done in wireless sensor networks (where it
also matters to reduce the number of messages, but for other reasons:
power consumption), we can do aggregation in nodes. In difference to
traditional distributed systems, this works, because it is easy to
deploy custom software on every node in the network. Classical
distributed systems do not typically allow this. Furthermore, reducing
the number of messages at the price of higher complexity does not make
sense there.

@@

\begin{itemize}
\item Arguing that group communication/atomic broadcast and showing
  that we can do it well might be enough (appeal to atomic broadcasts
  as the foundation for other distributed algorithms, and the MPI
  primitives, that can/need to be mapped to group-communication (have
  a list of them) 
\item Need higher-level applications, such as:
  \begin{itemize}
  \item octopus
    \begin{itemize}
    \item higher level application?
    \end{itemize}
  \item capability system? \stefan{Ask Simon how expensive revocation
      currently is}
  \item a database with replication and consistency maintenance
    \begin{itemize}
    \item SharedDB? (buffers between operators, but no synchronization
      except for access to buffers, which are multiple writers, one
      reader)
    \item Crescando (some kind of state machine replication)
    \end{itemize}
  \end{itemize}
\item Machine model for topology aware tree
  \begin{itemize}
  \item Talked to Pravin about this a bit
  \end{itemize}
\item MPI collectives (as in~\cite{Tu2008})
\end{itemize}

>>>>

Atomic broadcast is a building-block for many higher-level distributed
algorithm such as state-machine replication. Such distributed
algorithms are increasingly used to overcome scalability challenges on
multicore machines. Examples are databases~\cite{Salomie2011,
  Wiesmann2000} and operating systems~\cite{fos:osr09, tornado:osdi99,
  barrelfish:sosp09}.

@@

@@

A multicore aware broadcast has been implemented in~\cite{Tu2008}. It
takes the cache architecture of the machine (and only one particular?)
into consideration for building the broadcast. It is however for
larger messages (starting at 4K) and it does not seem to be performing
very well (numbers are given in microseconds -6). 4K messages are 52
microseconds, which is 150000 cycles. For synchronization primitives,
a cache line is sufficient.

@@

\subsection{Aggregation for convergecast}

Convergecast is a way of collecting information from nodes (i.e.\ it
is kind of the reverse of a broadcast). On the return path,
information can be aggregated. One example is batching, where several
pieces of information are grouped in the same message to reduce packet
processing overhead. Another example is aggregation, i.e.\
pre-processing of data. One example is pre-calculating the average
value of child messages if the sink node is not interested in
individual values.

Similarly to what has been done in wireless sensor networks (where it
also matters to reduce the number of messages, but for other reasons:
power consumption), we can do aggregation in nodes. In difference to
traditional distributed systems, this works, because it is easy to
deploy custom software on every node in the network. Classical
distributed systems do not typically allow this. Furthermore, reducing
the number of messages at the price of higher complexity does not make
sense in classical systems.

@@

\begin{itemize}
\item Arguing that group communication/atomic broadcast and showing
  that we can do it well might be enough (appeal to atomic broadcasts
  as the foundation for other distributed algorithms, and the MPI
  primitives, that can/need to be mapped to group-communication (have
  a list of them) 
\item Need higher-level applications, such as:
  \begin{itemize}
  \item capability system? \stefan{Ask Simon how expensive revocation
      currently is}
  \item a database with replication and consistency maintenance
    \begin{itemize}
    \item SharedDB? (buffers between operators, but no synchronization
      except for access to buffers, which are multiple writers, one
      reader)
    \item Crescando (some kind of state machine replication)
    \end{itemize}
  \end{itemize}
\item Machine model for topology aware tree
  \begin{itemize}
  \item Talked to Pravin about this a bit
  \end{itemize}
\item MPI collectives (as in~\cite{Tu2008})
\end{itemize}

\subsection{Octopus}

One idea is to make Octopus a decentralized application. Octopus is
a coordination service implemented for the Barrelfish Multikernel
operating system. It is, however, centralized.

The idea is to extend Octopus such that it can be replicated within a
multicore machine. This will retain scalability for future manycore
machines and potentially outperform a centralized version of Octopus
especially in contented systems.

One foundation block of Octopus is a key-value store. Since there is
only one copy, the data in there is always consistent. If replicated,
some sort of consistency guarantee would have to be implemented. One
example is \emph{linearizability}, i.e.\ ``accesses occur one at a time
in, some sequential order that is consistent with the order of
invocations and responses''~\cite{lynch}.

The question is whether a key-value store as a backend for
synchronization primitives even makes sense.

One concrete idea is to have one copy of the Octopus data-store
replicated on every NUMA node. One core on every node acts as
\emph{coordinator}, we call the other nodes \emph{slaves}.
Coordinators communicate on a ring. Communication on a local NUMA node
can be realized using read-only shared memory for read access. Write
access needs to be globally ordered and hence communication with other
replicas is required.

The idea of having a ring is somewhat related to SharedDB. It
optimizes for throughput, not latency!

\begin{description}
\item[Updates on global state] Slaves direct their updates to the
  local coordinator. Coordinators send the update request $r$ along
  the ring and wait for them to be send back. While passing on
  requests, coordinators can set a flag to indicate conflicts. A
  conflict is another colliding request $r_o$ that is already in
  flight. I think we will have to send activity vectors along with
  this information! Otherwise, the coordinators requesting $r$ and
  $r_o$ will both prevent the other thing from happening and none of
  them will be applied! Maybe we need a two-phase commit.
\item[Barriers] The coordinator knows how many of its slaves need to
  enter the barrier. It can then just aggregate local barrier\_enters
  calls and only communicate this information to the others once all
  children are ready. 
\end{description}

\subsubsection{Load balancing}

We could build several rings, that do not always include all
coordinators and work with read- and write-sets. We can select these
rings such that they nicely integrate with the physical topology. They
can also have different sizes (e.g.\ a smaller one for read
operations and larger ones for writes).
